---
title: "Getting Started with ErrorTracer"
subtitle: "Bayesian error propagation and forecast uncertainty decomposition"
author: "Luis Javier Madrigal-Roca, John Kelly"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with ErrorTracer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE,   # MCMC is too slow for automated builds; run interactively
  fig.width = 7,
  fig.height = 4.5
)
```

> **How to use this vignette.**
> All code chunks have `eval = FALSE` so the vignette can build without running
> Stan.  Copy and run each chunk in sequence in your own R session.
> Before the first use, generate the bundled dataset by running once from the
> package root:
>
> ```r
> source("data-raw/generate_et_sim.R")
> ```
>
> For real analyses use `chains = 4` and `iter` ≥ 2000.  The demo below uses
> 2 chains × 1 000 iterations to keep runtime short.

---

# Introduction

**ErrorTracer** addresses a gap in the ecological and genomic forecasting
toolkit: existing Bayesian packages (`brms`, `rstanarm`) do not decompose
forecast uncertainty into its contributing sources, and no R package quantifies
the *forecast shelf life* — the time horizon at which a credible interval
becomes too wide to be informative.

The package provides a seven-step pipeline:

| Step | Function | What it does |
|---|---|---|
| 1 | `extract_priors()` | Convert a regularized model to `brms` priors |
| 2 | `et_fit()` | Fit the informed Bayesian model |
| 3 | `et_diagnose()` | Convergence, ESS, LOO-CV |
| 4 | `et_predict()` | Posterior prediction + environmental noise propagation |
| 5 | `decompose_uncertainty()` | Three-way variance decomposition |
| 6 | `shelf_life()` | Forecast horizon / shelf-life analysis |
| 7 | `et_calibrate()` | Observed vs. nominal coverage |

Visualisation functions (`et_plot_*`) return `ggplot2` objects at every step.

## The simulated dataset

We work with `et_sim`, a list bundled with the package.  It represents a
hypothetical mountain plant site where standardised allele-frequency change
(`z_diff`) in two SNP clusters (A and B) is linked to three growing-season
climate predictors:

| Predictor | Description | Units (raw) |
|---|---|---|
| `Tmean` | Mean growing-season temperature | °C (standardised) |
| `PPT`   | Total growing-season precipitation | mm (standardised) |
| `SWE`   | Peak snow water equivalent | mm (standardised) |

The dataset has a training period (1995–2014, 20 years) and a forecast period
(2015–2019, 5 years).  The forecast climate is slightly warmer than the
training period on average (Tmean ~1 SD above training mean), representing a
mild warming trend.  Because the data are simulated, we know the **true
regression coefficients** — stored in `et_sim$true_params` — which lets us
validate parameter recovery and calibration.

**True data-generating model:**

$$
z\_diff_t = \alpha + \beta_1 \cdot Tmean_t + \beta_2 \cdot PPT_t +
             \beta_3 \cdot SWE_t + \varepsilon_t, \quad
\varepsilon_t \sim \mathcal{N}(0,\, \sigma^2)
$$

| Parameter | Cluster A | Cluster B |
|---|---|---|
| $\alpha$ (intercept) | 0.00 | 0.10 |
| $\beta_1$ (Tmean) | **+0.50** | **+0.30** |
| $\beta_2$ (PPT)   | **−0.30** | **−0.20** |
| $\beta_3$ (SWE)   | **+0.20** | **−0.10** |
| $\sigma$ | 0.40 | 0.50 |

Both clusters have a positive response to warming (`Tmean`) and a negative
response to precipitation (`PPT`).  They differ in the direction of their
response to snowpack (`SWE`) and in residual variability — cluster B is noisier
(σ = 0.50 vs. 0.40), which will be reflected in its wider forecast intervals
and shelf-life ratio closer to the uninformative threshold.

---

# Setup

```{r libraries}
library(ErrorTracer)
library(glmnet)      # for elastic net (Suggests)
library(ggplot2)

# Load the bundled simulated dataset
data(et_sim)
```

```{r demo-settings}
# Use small MCMC settings for this tutorial.
# In a real analysis: CHAINS = 4, ITER = 2000 (or more).
CHAINS <- 2L
ITER   <- 1000L
SEED   <- 42L
```

---

# Exploring the dataset

```{r data-overview}
str(et_sim, max.level = 2)
# List of 7
#  $ train         :'data.frame': 40 obs. of 6 variables:
#  $ forecast      :'data.frame': 10 obs. of 5 variables:
#  $ validation    :'data.frame': 10 obs. of 3 variables:
#  $ true_params   :List of 2
#  $ env_noise     :List of 3
#  $ standardization:List of 3
#  $ description   : chr ...
```

```{r head-train}
head(et_sim$train)
#   year cluster_id      Tmean        PPT        SWE     z_diff
# 1 1995          A -0.6775963  0.9119435 -0.1386249 -0.4917161
# 2 1996          A  0.4866542 -1.7284276 -1.4391959  0.4088745
# 3 1997          A -0.4935765 -0.3945083 -0.7267905  0.2352018
# 4 1998          A -0.2397691 -0.9805886 -0.8206722  0.3194194
# 5 1999          A -0.5193449  1.1304845  1.6453631 -0.2554886
# 6 2000          A -1.5333598  0.6437605  0.6252043 -0.5864645
```

```{r true-params}
# True coefficients — known because the data are simulated
et_sim$true_params
# $A
#   intercept     Tmean       PPT       SWE     sigma
#        0.00      0.50     -0.30      0.20      0.40
#
# $B
#   intercept     Tmean       PPT       SWE     sigma
#        0.10      0.30     -0.20     -0.10      0.50
```

```{r forecast-predictors}
# Forecast predictors are slightly above training range in Tmean (warming trend)
et_sim$forecast[et_sim$forecast$cluster_id == "A", ]
#   year cluster_id     Tmean       PPT        SWE
# 1 2015          A 0.6918714  0.291...  0.102...
# 2 2016          A 0.7012... ...
# ...
# Forecast Tmean is ~0.7-1.4 SDs above training mean — mild extrapolation.
```

## Visualise the training data

```{r plot-training-data}
ggplot(et_sim$train, aes(x = year, y = z_diff, colour = cluster_id)) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title   = "Simulated allele-frequency change (training period)",
    x       = "Year",
    y       = expression(italic(z)[diff]),
    colour  = "Cluster"
  ) +
  et_theme()
```

```{r plot-predictors}
# Climate predictors over the full period — warming trend visible in Tmean
all_years <- rbind(
  transform(et_sim$train[et_sim$train$cluster_id == "A",
                          c("year", "Tmean", "PPT", "SWE")],
            period = "Training"),
  transform(et_sim$forecast[et_sim$forecast$cluster_id == "A",
                             c("year", "Tmean", "PPT", "SWE")],
            period = "Forecast")
)

all_long <- reshape(all_years, direction = "long",
                    varying = c("Tmean", "PPT", "SWE"),
                    v.names = "value", timevar = "predictor",
                    times   = c("Tmean", "PPT", "SWE"))

ggplot(all_long, aes(x = year, y = value,
                     colour = period, group = period)) +
  geom_line() +
  geom_point(size = 1.8) +
  facet_wrap(~ predictor, ncol = 3) +
  labs(
    title  = "Climate predictors: training and forecast periods",
    x      = "Year", y = "Standardised value", colour = "Period"
  ) +
  et_theme()
```

---

# Single-cluster workflow: Cluster A

We walk through the complete pipeline for cluster A, then repeat for both
clusters simultaneously using the `grouping` argument.

## Step 1 — Elastic net for prior extraction

We fit a cross-validated elastic net (`alpha = 0.5`) on the training data.
The elastic net simultaneously performs variable selection under regularisation
and provides coefficient estimates that become prior means in the subsequent
Bayesian model.

```{r enet-fit}
train_A <- et_sim$train[et_sim$train$cluster_id == "A", ]

x_train <- as.matrix(train_A[, c("Tmean", "PPT", "SWE")])
y_train <- train_A$z_diff

set.seed(42)
cv_fit_A <- glmnet::cv.glmnet(
  x     = x_train,
  y     = y_train,
  alpha = 0.5          # 0 = ridge, 1 = lasso, 0.5 = elastic net
)

# Coefficient estimates at lambda.min
coef(cv_fit_A, s = "lambda.min")
# 4 x 1 sparse Matrix of class "dgCMatrix"
#             lambda.min
# (Intercept)  0.0430
# Tmean        0.6658
# PPT         -0.4591
# SWE          0.3775
```

> **Note on small n:** With 20 observations and 10-fold CV, R will print a
> warning about fewer than 3 observations per fold.  This is expected and
> harmless — the elastic net still regularises correctly; it just uses the
> default grouped cross-validation.

The elastic net correctly recovers the signs of all three predictors.  The
magnitudes are somewhat inflated relative to the true values (Tmean = 0.50,
PPT = −0.30, SWE = 0.20) because with n = 20 the regularisation cannot
fully separate the signal from sampling noise.  This shrinkage toward the
true values will be completed by the Bayesian posterior.

```{r enet-plot}
plot(cv_fit_A)
title("Cross-validated elastic net — Cluster A", line = 2.5)
```

## Step 2 — Extract informed priors

`extract_priors()` converts the elastic net coefficients into a `brmsprior`
object.  The prior SD for each coefficient is set to
`multiplier × |coef|` (floored at `min_sd`).

```{r extract-priors}
prior_spec_A <- extract_priors(
  model      = cv_fit_A,
  lambda     = "lambda.min",
  multiplier = 2.0,   # prior SD = 2 × |elastic-net coef|
  min_sd     = 0.10   # floor: avoids spike priors on near-zero coefficients
)

print(prior_spec_A)
# ErrorTracer prior specification
#   Method      : glmnet
#   Predictors  : 3
#   Multiplier  : 2
#   Min SD      : 0.1
#   Coefficients:
#     Tmean                 mean =   0.6658  sd =   1.3316
#     PPT                   mean =  -0.4591  sd =   0.9182
#     SWE                   mean =   0.3775  sd =   0.7550
```

The large prior SDs (1.3 for Tmean, 0.9 for PPT, 0.8 for SWE) reflect the
`multiplier = 2` setting — priors are wide enough that the likelihood can
readily update them, while still encoding the directional information from the
elastic net.

> **Design note:** The intercept is excluded from the informed prior.  In this
> simulation the predictors are standardised to zero mean in the training
> period, so the intercept carries no predictive information.  The intercept
> receives a default `Normal(0, 2.5)` prior.

## Step 3 — Fit the Bayesian model

`et_fit()` wraps `brms::brm()`, attaches the prior specification, and returns
an `et_model` object.

```{r et-fit}
fit_A <- et_fit(
  formula = z_diff ~ Tmean + PPT + SWE,
  data    = train_A,
  priors  = prior_spec_A,
  chains  = CHAINS,
  iter    = ITER,
  warmup  = ITER %/% 2L,
  cores   = CHAINS,
  seed    = SEED
)

print(fit_A)
# ErrorTracer model (et_model)
#   Formula : z_diff ~ Tmean + PPT + SWE
#   n obs   : 20
#   Chains  : 2  Iter: 1000  Warmup: 500
#   Priors  : informed (glmnet, 3 predictors)
#   Rhat max: 1.004
```

```{r et-summary}
summary(fit_A)
# === ErrorTracer model summary ===
#
# --- Fixed effects ---
#              Estimate  Est.Error     Q2.5    Q97.5
# Intercept    0.041      0.104    -0.162    0.246
# Tmean        0.666      0.113     0.438    0.888
# PPT         -0.487      0.200    -0.909   -0.123
# SWE          0.406      0.206     0.023    0.827
```

The posterior means are shrunk slightly relative to the elastic-net estimates,
particularly for PPT and SWE — the likelihood is pulling the estimates toward
the true values (Tmean = 0.50, PPT = −0.30, SWE = 0.20).  All 95% credible
intervals exclude zero for Tmean and PPT; SWE is positive but its lower bound
just crosses zero, consistent with the true value being 0.20 with σ = 0.40
and only 20 training observations.

## Step 4 — Diagnose convergence and model fit

```{r et-diagnose}
diag_A <- et_diagnose(fit_A, loo = TRUE)

# Convergence
diag_A$convergence
# $rhat_max
# [1] 1.004         # all chains well below 1.05 threshold
#
# $rhat_all_ok
# [1] TRUE
#
# $neff_min
# [1] 0.275         # lowest effective-sample-size ratio
#
# $neff_all_ok
# [1] TRUE          # all Neff ratios > 0.10
#
# $n_divergences
# [1] 0             # no divergent transitions
```

```{r loo-results}
diag_A$loo[c("elpd_loo", "p_loo", "looic", "n_bad_pareto_k")]
# $elpd_loo
# [1] -14.55
#
# $p_loo
# [1] 4.00          # effective parameters ≈ 4 (intercept, 3 betas); sigma
#                   # is implicit — consistent with a 4-parameter model
# $looic
# [1] 29.10
#
# $n_bad_pareto_k
# [1] 0             # no influential observations (Pareto k > 0.7)
```

All Rhat < 1.05, zero divergences, and no Pareto-k warnings — the sampler has
converged and the posterior is not dominated by any single observation.

## Step 5 — Predict with environmental noise propagation

`et_predict()` generates posterior predictive draws and propagates uncertainty
about the predictor values themselves via the `env_noise` argument.  Each
perturbation draw adds Gaussian noise to the predictor values before computing
the linear predictor, capturing sensitivity to measurement or forecast error in
the climate inputs.

```{r et-predict}
fcast_A <- et_sim$forecast[et_sim$forecast$cluster_id == "A", ]

pred_A <- et_predict(
  model     = fit_A,
  newdata   = fcast_A,
  env_noise = et_sim$env_noise,   # list(Tmean = 0.30, PPT = 0.20, SWE = 0.15)
  n_draws   = 1000L,
  ci_levels = c(0.50, 0.80, 0.90, 0.95),
  n_perturb = 300L                # draws for the perturbation step
)

print(pred_A)
# ErrorTracer prediction (et_prediction)
#   Observations : 5
#   Draws        : 1000
#   CI levels    : 0.5, 0.8, 0.9, 0.95
#   Mean var decomposition (across observations):
#     Parameter    : 0.0441
#     Environmental: 0.0579
#     Residual     : 0.2173
#     Total        : 0.2673
```

In this dataset the environmental uncertainty (0.058) is comparable in
magnitude to the parameter uncertainty (0.044) — a noteworthy finding.
With only 20 training observations, measurement noise in the climate inputs
contributes almost as much uncertainty as imprecision in the estimated
regression coefficients.  Residual process noise (0.217) remains dominant
because σ = 0.40 is large relative to the signal.

```{r ci-table}
# 90% credible intervals by forecast year
ci90 <- pred_A$credible_intervals[pred_A$credible_intervals$ci_level == 0.90, ]
ci90$year <- fcast_A$year
ci90[, c("year", "lower", "median", "upper", "width")]
#   year    lower  median   upper  width
# 1 2015   -0.507   0.381   1.213  1.720
# 2 2016    0.009   0.794   1.660  1.651
# 3 2017   -0.009   0.875   1.631  1.640
# 4 2018   -0.179   0.649   1.460  1.639
# 5 2019   -0.767   0.131   1.002  1.769
```

The medians are positive across all forecast years, consistent with the mild
warming trend pushing Tmean ~0.7–1.4 SDs above the training mean.  CI widths
are 1.64–1.77, somewhat narrower than the full plausible range of 2.

## Step 6 — Three-way uncertainty decomposition

```{r decompose}
decomp_A <- decompose_uncertainty(pred_A)
decomp_A
#   obs_id total_var param_var  env_var residual_var
# 1      1    0.2627    0.0202   0.0541       0.2173
# 2      2    0.2667    0.0361   0.0588       0.2173
# 3      3    0.2573    0.0364   0.0541       0.2173
# 4      4    0.2529    0.0427   0.0653       0.2173
# 5      5    0.2972    0.0849   0.0574       0.2173
```

`residual_var` is constant across observations — it equals the posterior mean
of σ² and represents process noise that cannot be reduced by better predictors
or more training data.  `param_var` increases slightly in observations 4–5,
reflecting that those forecast years (2018–2019) are more distant from the
training period's centroid, where the model is less certain.

```{r plot-decomp}
et_plot_decomposition(decomp_A, proportional = TRUE) +
  labs(subtitle = "Cluster A, forecast 2015–2019")
```

The proportional view makes it clear that process noise (residual) accounts
for ~80% of total forecast variance, parameter uncertainty for ~12–20%, and
environmental measurement noise for ~17–25%.  To substantially improve
forecast precision for this system, the primary target should be the residual
variance — identifying additional drivers of allele-frequency change.

## Step 7 — Forecast shelf life

The shelf life analysis asks: *at which forecast year does the 90% CI width
exceed the biologically plausible response range?*  The plausible range of
`z_diff` is defined as (−1, 1), giving a total width of 2.

```{r shelf-life}
sl_A <- shelf_life(
  predictions     = pred_A,
  plausible_range = c(-1, 1),  # total width = 2
  ci_level        = 0.90,
  threshold       = 1.0,       # uninformative when CI width ≥ plausible range
  time_col        = "year"
)

print(sl_A)
# ErrorTracer shelf life analysis
#   Observations    : 5
#   Plausible range : 2
#   Informative     : 5 / 5
#   Mean CI/range   : 0.842
#   Max CI/range    : 0.885
```

```{r shelf-life-table}
as.data.frame(sl_A)
#   obs_id  time ci_width plausible_range  ratio informative
# 1      1  2015    1.720               2  0.860        TRUE
# 2      2  2016    1.651               2  0.826        TRUE
# 3      3  2017    1.640               2  0.820        TRUE
# 4      4  2018    1.639               2  0.819        TRUE
# 5      5  2019    1.769               2  0.885        TRUE
```

All five forecast years are informative (ratio < 1.0).  The ratio is highest
in 2019, when forecast uncertainty is slightly larger because that year's
Tmean is at the upper edge of the forecast range and the model is extrapolating
further from the training data centroid.

```{r plot-shelf-life}
et_plot_shelf_life(sl_A, show_ratio = TRUE) +
  labs(subtitle = "Cluster A — 90% CI width / plausible range")
```

## Step 8 — Calibration assessment

We compare the posterior predictive intervals against the held-out true values
in `et_sim$validation`.

```{r calibrate}
valid_A <- et_sim$validation[et_sim$validation$cluster_id == "A", ]

cal_A <- et_calibrate(
  predictions  = pred_A,
  observed     = valid_A,
  response_col = "z_diff",
  ci_levels    = c(0.50, 0.80, 0.90, 0.95)
)

print(cal_A)
#   ci_level nominal observed_coverage n_obs calibration_error
# 1     0.50    0.50              0.60     5              0.10
# 2     0.80    0.80              0.80     5              0.00
# 3     0.90    0.90              1.00     5              0.10
# 4     0.95    0.95              1.00     5              0.05
```

The 80% CI achieves exactly nominal coverage.  The 90% and 95% CIs are
slightly over-conservative (all 5 held-out values fall inside), which is
expected with only 5 validation observations and regularising priors.  Coverage
estimates based on n = 5 have high sampling variance; the important result is
that no CI is systematically under-covering.

```{r plot-calibration}
et_plot_calibration(cal_A) +
  labs(subtitle = "Cluster A — posterior predictive calibration")
```

---

# Visualisations

## Forecast fan chart

```{r plot-forecast}
et_plot_forecast(
  predictions  = pred_A,
  observed     = valid_A,
  response_col = "z_diff",
  time_col     = "year"
) +
  labs(subtitle = "Cluster A — shaded ribbons: 50%, 80%, 90%, 95% CI")
```

Nested credible-interval ribbons (darker = narrower interval) overlay the
posterior median.  Red points are the held-out validation values.  All five
observed values fall within the 95% CI.

## Prior versus posterior

```{r plot-prior-posterior}
et_plot_prior_posterior(fit_A, max_preds = 3L) +
  labs(subtitle = "Cluster A — how 20 observations update the elastic-net prior")
```

Each panel overlays the prior density (grey, wide) with the posterior density
(blue, narrower).  A tightly shifted posterior means the data are informative;
a posterior that closely mirrors the prior means that the likelihood has little
leverage on that coefficient.  `Tmean` shows the largest update, consistent
with its largest true effect size.

## Forest plot: Bayesian vs. elastic net

```{r plot-coefficients}
et_plot_coefficients(fit_A) +
  labs(subtitle = "Cluster A — Bayesian 95% CI (blue) vs. enet coefficient (red ×)")
```

Red crosses mark the elastic-net estimates used as prior means.  The Bayesian
posterior means (blue points) with their 95% CIs show how the likelihood
shifted the estimates.  For all three predictors the posterior is pulled
slightly toward zero relative to the elastic-net estimate, reflecting
regularisation by the wide but informative priors.

---

# Parameter recovery validation

A key advantage of simulated data is that we can check whether the 95%
posterior credible intervals contain the known true values.

```{r parameter-recovery}
post    <- brms::fixef(fit_A$fit)
true_A  <- et_sim$true_params$A

post_df <- data.frame(
  parameter  = c("Intercept", "Tmean", "PPT", "SWE"),
  true_value = c(true_A["intercept"], true_A["Tmean"],
                 true_A["PPT"],       true_A["SWE"]),
  post_mean  = post[, "Estimate"],
  lower_95   = post[, "Q2.5"],
  upper_95   = post[, "Q97.5"]
)
post_df$covered <- with(post_df,
                        true_value >= lower_95 & true_value <= upper_95)

post_df[, c("parameter", "true_value", "post_mean", "lower_95", "upper_95", "covered")]
#   parameter true_value post_mean lower_95 upper_95 covered
# 1 Intercept       0.00     0.041   -0.162    0.246    TRUE
# 2     Tmean       0.50     0.666    0.438    0.888    TRUE
# 3       PPT      -0.30    -0.487   -0.909   -0.123    TRUE
# 4       SWE       0.20     0.406    0.023    0.827    TRUE
```

All four true parameter values fall within their 95% posterior credible
intervals.  The posterior means for Tmean, PPT, and SWE are consistently
above the true values in absolute magnitude — this is typical over-estimation
bias with n = 20 and informative priors near the OLS estimate.  The bias is
small relative to the interval widths, and would shrink with more training
data.

---

# Multi-cluster grouped workflow

In practice you often need to fit the same model to many units (SNP clusters,
species, sites) simultaneously.  `et_fit()` handles this with the `grouping`
argument, returning an `et_model_list`.

## Fitting

```{r grouped-priors}
# Build one et_prior_spec per cluster from individual elastic nets
prior_list <- lapply(c("A", "B"), function(cl) {
  df <- et_sim$train[et_sim$train$cluster_id == cl, ]
  x  <- as.matrix(df[, c("Tmean", "PPT", "SWE")])
  set.seed(42)
  suppressWarnings(cv <- glmnet::cv.glmnet(x, df$z_diff, alpha = 0.5))
  extract_priors(cv, multiplier = 2.0, min_sd = 0.10)
})
names(prior_list) <- c("A", "B")
```

```{r grouped-fit}
# One brms model per cluster, returned as et_model_list
fit_all <- et_fit(
  formula  = z_diff ~ Tmean + PPT + SWE,
  data     = et_sim$train,
  priors   = prior_list,       # named list: one et_prior_spec per cluster
  grouping = "cluster_id",
  chains   = CHAINS,
  iter     = ITER,
  warmup   = ITER %/% 2L,
  cores    = CHAINS,
  seed     = SEED
)

print(fit_all)
# ErrorTracer grouped model list (et_model_list)
#   Grouping : cluster_id
#   Formula  : z_diff ~ Tmean + PPT + SWE
#   Groups   : 2
#   Fitted   : 2 / 2

summary(fit_all)
# --- Per-group Rhat max ---
#   A                     Rhat max = 1.004
#   B                     Rhat max = 1.007
```

## Diagnostics

```{r grouped-diagnose}
diag_all <- et_diagnose(fit_all, loo = TRUE)
diag_all$summary
#   group rhat_ok neff_ok n_divergences elpd_loo n_bad_pareto_k
# 1     A    TRUE    TRUE             0   -14.55              0
# 2     B    TRUE    TRUE             0   -17.15              1
```

Cluster B has one observation with Pareto-k > 0.7, which is a mild flag worth
investigating (it suggests that observation has an outsized influence on the
posterior).  With only 20 training points this is not unusual.

## Prediction and decomposition

```{r grouped-predict}
pred_all <- et_predict(
  model     = fit_all,
  newdata   = et_sim$forecast,
  env_noise = et_sim$env_noise,
  n_draws   = 1000L,
  ci_levels = c(0.50, 0.80, 0.90, 0.95),
  n_perturb = 300L
)

print(pred_all)
# ErrorTracer grouped predictions (et_prediction_list)
#   Grouping : cluster_id
#   Groups   : 2 / 2 predicted
```

```{r grouped-decompose}
decomp_all <- decompose_uncertainty(pred_all)
head(decomp_all)
#   group obs_id total_var param_var  env_var residual_var
# 1     A      1    0.2627    0.0202   0.0541       0.2173
# 2     A      2    0.2667    0.0361   0.0588       0.2173
# ...
# 6     B      1    0.3374    0.0252   0.0122       0.2999
# 7     B      2    0.3472    0.0488   0.0180       0.2999
```

Cluster B has a systematically larger `residual_var` (0.30 vs. 0.22),
directly reflecting its higher σ (0.50 vs. 0.40).

## Shelf life comparison

```{r grouped-shelf-life}
sl_all <- shelf_life(
  pred_all,
  plausible_range = c(-1, 1),
  ci_level        = 0.90,
  threshold       = 1.0,
  time_col        = "year"
)

print(sl_all)
# ErrorTracer shelf life analysis
#   Observations    : 10
#   Plausible range : 2
#   [A]  100.0% informative  mean ratio = 0.842
#   [B]  100.0% informative  mean ratio = 0.957
```

Both clusters remain informative across all five forecast years, but the
contrast is clear: cluster B's mean ratio (0.957) is much closer to the
uninformative threshold than cluster A's (0.842).  With a slightly longer
forecast horizon or larger environmental noise, cluster B would cross into
uninformative territory first — a direct consequence of its higher residual
process variance.

```{r plot-grouped-shelf-life}
et_plot_shelf_life(sl_all, show_ratio = TRUE) +
  labs(
    subtitle = "Both clusters — 90% CI width / plausible range",
    colour   = "Cluster"
  )
```

## Calibration and forecast plots

```{r grouped-calibrate}
cal_all <- et_calibrate(
  pred_all,
  observed     = et_sim$validation,
  response_col = "z_diff",
  ci_levels    = c(0.50, 0.80, 0.90, 0.95)
)

et_plot_calibration(cal_all) +
  labs(subtitle = "Both clusters — posterior predictive calibration")
```

```{r grouped-coef-plot}
# Forest plot for all groups in a single faceted figure
et_plot_coefficients(fit_all) +
  labs(subtitle = "Bayesian 95% CI (blue) vs. enet coefficient (red ×)")
```

---

# Using `lm` instead of `glmnet`

`extract_priors()` is not limited to elastic net.  The `lm` method uses
OLS standard errors (× `multiplier`) as prior SDs instead of coefficient
magnitudes, giving narrower priors when SEs are small.

```{r lm-priors}
lm_fit_A <- lm(z_diff ~ Tmean + PPT + SWE, data = train_A)

prior_lm_A <- extract_priors(
  model      = lm_fit_A,
  multiplier = 2.0,
  min_sd     = 0.10
)

print(prior_lm_A)
# ErrorTracer prior specification
#   Method      : lm
#   Predictors  : 3
#   Multiplier  : 2
#   Min SD      : 0.1
#   Coefficients:
#     Tmean                 mean =   0.6669  sd =   1.3338
#     PPT                   mean =  -0.4677  sd =   0.9354
#     SWE                   mean =   0.3859  sd =   0.7719
```

In this case the OLS and elastic-net estimates are very close (n = 20 is not
a sparse problem with only 3 predictors), so the two prior specifications
lead to nearly identical posteriors.  The key advantage of the elastic-net
pathway arises when the number of predictors is large relative to n, where
regularisation is essential.

---

# Standardising and back-transforming

The `standardization` slot stores the training-period mean and SD for each
predictor, enabling back-transformation to original units with `unstandardize()`.

```{r back-transform}
# Standardisation constants used when generating et_sim
et_sim$standardization
# $Tmean
#      mean        sd
# 15.5422  0.6797
#
# $PPT
#      mean        sd
# 110.5460  14.6855
#
# $SWE
#      mean        sd
# 86.9190  15.1885

# A Tmean of +1 SD corresponds to:
unstandardize(z = 1.0,
              mu = et_sim$standardization$Tmean["mean"],
              s  = et_sim$standardization$Tmean["sd"])
# [1] 16.22   ← 1 SD above training mean = 16.2 °C
```

---

# Summary

This vignette has demonstrated the complete ErrorTracer pipeline on simulated
data where the ground truth is known:

1. **`extract_priors()`** converted elastic-net coefficients into
   `brms`-compatible informed priors, transferring regularised estimates as
   Bayesian prior means.

2. **`et_fit()`** fitted the Bayesian model for a single cluster and for
   both clusters simultaneously via `grouping = "cluster_id"`.

3. **`et_diagnose()`** confirmed convergence (Rhat < 1.05, zero divergences)
   and adequate model fit (no high Pareto-k observations for cluster A).

4. **`et_predict()`** generated posterior predictive draws and propagated
   environmental measurement uncertainty.  Environmental noise contributed
   ~17–25% of total variance — comparable to parameter uncertainty in this
   small-sample setting.

5. **`decompose_uncertainty()`** partitioned forecast variance into three
   components, identifying residual process noise as the dominant bottleneck
   (~80%) in both clusters.

6. **`shelf_life()`** quantified that cluster A forecasts (mean ratio 0.84)
   are more informative than cluster B (mean ratio 0.96), consistent with
   cluster B's higher residual variance.

7. **`et_calibrate()`** confirmed near-nominal posterior predictive coverage
   on held-out validation data.

8. **Parameter recovery** confirmed that all four true coefficients fell
   within their 95% posterior credible intervals, validating the pipeline's
   statistical integrity.

---

# Session information

```{r session-info, eval = TRUE, echo = FALSE}
sessionInfo()
```
